---
title: "GSS challange"
author: "Kaylie Brehm!"
date: "Summer 2022"
output: html_document
---







```{r include=FALSE,echo=FALSE}
source(file="fisherXY.R")
require(tidyverse)
require(tigerstats)
require(knitr)
require(foreign)
GSS <-read.dta(file="GSS2018.dta")
GSS <- droplevels(GSS)
mydata <- select(GSS,race,cappun)
mydata <- na.omit(mydata)
```

# Introduction

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?.  If you skimp in this section expect to be marked down in your grade. Paint a picture of the project a couple of paragraphs is expected!-->

Is attitude toward capitol punishment (cappun) independent of race (race)?
If not, then whose view differs from whom??

$H_0$: Attitude towards capital punishment is not related to race.

$H_a$: Attitude towards capital punishment is related to race.

# Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method. Discuss what data manipulations may be needed for example subsetting or other binning procedures and what sort of graphical displays and inferential tools will be used!-->

Both race and capital punishment are categorical variables so the analysis technique is CAT~CAT; plot is a bar plot, summary is a cross table, and inference is a chi-square test. We finally use a fisher x-y test to analyze whose views differ from whom.

# Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

## Descriptive Results

### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question. Bottom line is to explain the features of the graph that are relevant to conforming or departing to what is expected by your null hypothesis !-->




```{r}
dd2 <- mydata %>% group_by(race,cappun) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=race,y=count,fill=cappun))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=race,y=prcnt*100,fill=cappun)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```

Based on the data, it is apparent that the white race favor capital punishment far more than the black or other race. 

### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->


```{r}
table2 <- xtabs(~race + cappun, data=mydata)
rowPerc(table2)
colPerc(table2)
```

The top data set shows percentages for each race. About 68% of white people favor capital punishment. In the black race, about 46% favor capital punishment. In the other race, we see that about 56% favor capital punishment. We see the least variability in opinion from the black race and the most in the white.

## Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

`


```{r}
chisq.test(table2)
chisqtestGC(table2)
fisher.test(table2)
#fisherXY(mydata$race,mydata$cappun)
```

If there is a relationship between race and opinion, then there is a difference, meaning it is not 50/50 equal results. The Chi-Square adds up this difference and subtracts what we would expect if the null hypothesis were true. The P-Value is the probability that the null hypothesis is true. The p-value of this study is 0.0000000000000127. Since this p-value is under 0.05, I reject the null hypothesis due to it being so small.

```{r}
fisherXY(mydata$cappun,mydata$race)
```

When comparing the difference in opinion for the white and black race we see a p-value of 0.0000000000000575, so I reject the null hypothesis because it is under 0.05. The odds ratio means that white people are about 2.5 times more likely to be in favor of capital punishment than black people.


When comparing the difference in opinion for the white and other race we see a p-value of 0.0002819, so I reject the null hypothesis because it is under 0.05. The odds ratio means that white people are about 1.7 times more likely to be in favor of capital punishment than the "other" people.


When comparing the difference in opinion for the black and other race we see a p-value of 0.02479, so I reject the null hypothesis because it is under 0.05. The odds ratio means that black people are about 0.7 times more likely to be in favor of capital punishment than the "other" people.
 

# Discussion and Conclusion

<!--Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.!-->

The original question was "Is attitude toward capitol punishment (cappun) independent of race (race)?, If not, then whose view differs from whom??." The null hypothesis was "Attitude towards capital punishment is not related to race." I reject the null hypothesis due to a p-value so small, at 0.0000000000000127. We also observed p-values for each race compared to one another, testing their likelihood. Each of those p-values was also under 0.05, so I reject the null hypothesis on those as well. When observing difference we see the following: white people are about 2.5 times more likely to be in favor of capital punishment than black people, that white people are about 1.7 times more likely to be in favor of capital punishment than the "other" people, and that black people are about 0.7 times more likely to be in favor of capital punishment than the "other" people. With this observed difference, we do see a relationship between race and opinion on capital punishment. 
